A Simple Minibatch Acceptance Test for MCMC

### 2016-05-19 nips_2016.tex
Modified section 5.2 a bit

\author{
  Haoyu Chen \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{haoyuchen@berkeley.edu}
  \And
  Daniel Seita \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{seita@berkeley.edu}
  \And
  Xinlei Pan \\
  Graduate Program in Bioengineering \\
  UC Berkeley-UC San Francisco \\
  \texttt{xinleipan@berkeley.edu}
  \And 
  Biye Jiang \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{bjiang@berkeley.edu}
  \And
  John Canny \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{canny@berkeley.edu}
}

@@

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the mini-batch size; in the worst case, all $N$ data points may be needed in a single iteration.  A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties,
but comes at the cost of more computational time.

@@

In this section, we explore the convergence properties of our MH test. We first introduce some
notation. Let our true and approximated acceptance probabilities be $P_a = g(\Delta) = \Pr(\Delta +
X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} > 0)$.  Define $X_{\xi} = \varepsilon - X_{\rm norm}$
which means, invoking Equations~\ref{eq:relationship} and~\ref{eq:deconvolution}, that $X_\xi$ is
the random variable characterizing the accuracy of our $X_{\rm norm}$.  Thus, $\Delta' = \Delta +
X_{\rm norm} + X_\xi$. We assume the standard deviation of $X_{\rm norm}$ is less than
$\pi/\sqrt{3}$, so that we can use Equation~\ref{eq:deconvolution}. We denote $F_X(x)$ as the CDF of
$X$.

@@

\begin{proof}
See Appendix~\ref{app:theory1}.
\end{proof}

@@

\begin{proof}
See Appendix~\ref{app:theory2}.
\end{proof}

@@

\begin{proof}
See Appendix~\ref{app:theory3}.
\end{proof}

@@

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (capped at 600 for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in roughly 50 times more computation. For this particular experiment, that happened eight times.  In
contrast, our method keeps the minibatch size fixed at 200, but requires that ${\rm std}(\Delta') <
1.2$. Figure~\ref{fig:gauss_mix_2} (in the middle) shows a histogram of the estimated ${\rm
std}(\Delta')$ values each iteration. Only 75 iterations resulted in ${\rm std}(\Delta') \ge 1.2$
and for those, we skipped the rest of the iteration (setting $\theta_{t+1}=\theta_t$). To smooth the
${\rm std}(\Delta')$ values, we use moving average updates. The third plot in
Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta') \ge 1.2$. For six
minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with our MH test five times and the
number of violations averaged. As minibatch sizes increases, the number of violations decreases,
showing that our test is more applicable.

@@

In this experiment, we use logistic regression to classify the MNIST
dataset~\cite{lecun-mnisthandwrittendigit-2010}. For simplification, we only train the binary
classifier for digits 1 vs 7. The dataset consisted of 12007 training data and 1000 testing data.
Similarly, we use random walk as the proposer. We initilize the temperature as 3000, and decrease it
by $t_i = t_0 / (i + 1)^{0.5}$, where $i$ represent the $t^{th}$ iteration. Thus, the standard
derivation of $\Delta'$ is smaller than the threshold $\pi/\sqrt{3}$.  We compare our mini-batch MH
testing with the adaptive mini-batch MH testing. Figure 4 shows the log likelihood value and
prediction accuracy over training process. Since the adaptive MH testing method needs to increase
the mini-batch size adaptively, we use the cumulative number of data instance each algorithm
consumes as the horizontal axis. We find our mini-batch MH testing can achieve the similar (or even
better log likelihood/prediction accuracy) with consuming much fewer data points. Thus, we can
assert that our mini-batch MH testing is much more efficient than the adaptive one.

@@

We use this experiment to illustrate another property of our mini-batch MH testing: when the size of
mini-batch increase to a threshold, the accuracy / log likelihood of sampled data at a fix iteration
number will not improve with larger mini-batch size, see Fig.~4c. This is because the variance of
$\Delta'$ decrease with large mini-batch, and the error of our mini-batch MH testing is proportional
to our estimation error of variance of $\Delta'$. When the ${\rm Var}(\Delta')$ decrease to a
threshold, the error of our mini-batch MH testing can be ignored.

Figure~\ref{fig:logistic_fig} shows our results ... {\color{blue} Daniel: TODO finish description}

@@

In this experiment, we apply about mini-batch MH testing with SGHMC (Stochastic Gradient Hamiltonian
Monte Carlo~\cite{sghmc_2014}) proposer to sample the parameter instances from a fully connected
neural network. We use Higgs data set from University of California Irvine. It is a classification
problem to distinguish between a signal process which produces Higgs bosons and a background process
which does not. We build a 4 layers neural network to tackle this problem. The first layer is a
input layer, the last layer is a softmax layer, the intermediate layers are linear followed by
Sigmoid, norm, dropout, starting and ending with linear operation.

@@

\begin{itemize}[noitemsep]
    \item We use $N=10000$ samples of $x_i$.
    \item The covariance of the Gaussian random walk is ${\rm diag}(0.1, 0.1)$.
    \item We use 10000 iterations. One iteration involves one $\theta'$ proposal and a test to
    accept/reject.
    \item The mini-batch size is set at 100, so it is one percent of the total data. For the
    adaptive sampling approach, this is the amount that the minibatch size would keep increasing if
    we needed more precision. Our approach, of courses, keeps the size fixed.
    \item For adaptive sampling, the tolerance for deciding on a test is 0.05. The temperature for
    that is also set at $T=1$ because that algorithm was not designed to deal with temperatures.
    \item For our approach, to get an estimate of the standard deviation of $\Delta'$ (remember,
    $\Delta'$ is a minibatch approximation of $\Delta$) we take five other minibatches (randomly
    selected from the data) and take the standard deviation of those values.
\end{itemize}

@@

We applied our mini-batch Metropolis Hastings algorithm to a Bayesian logistic regression model, comparing with the adaptive sampling method. We tested both our method and the adaptive sampling method using a random walk proposer $q(\theta \mid \theta_t) = N (\theta_t, \sigma^2)$. Since the random walk proposer does not contain any information about the target distribution, it only relies on the Metropolis-Hastings test to converge to the correct distribution. Therefore, random walk proposer is the ideal choice for illustrating our algorithm and comparing with the adaptive sampling method.

>>>>

\author{
  Haoyu Chen \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{haoyuchen@berkeley.edu}
  \And
  Daniel Seita \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{seita@berkeley.edu}
  \And
  Xinlei Pan \\
  Department of Bioengineering \\
  University of California, Berkeley \\
  \texttt{xinleipan@berkeley.edu}
  \And 
  Biye Jiang \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{bjiang@berkeley.edu}
  \And
  John Canny \\
  Department of Computer Science \\
  University of California, Berkeley \\
  \texttt{canny@berkeley.edu}
}

@@

Unfortunately, computing $f$ requires the use of all $N$ training data points. Moreover, it is
difficult to design tests using substantially fewer than $N$ points that also satisfy detailed
balance. To reconcile these competing objectives,~\cite{cutting_mh_2014} proposes an adaptive
minibatch MCMC algorithm which uses a sequential hypothesis test. During each iteration, the
algorithm starts with a small minibatch of data and tests the hypothesis that the sample $\theta'$
should be accepted or rejected based on the likelihood ratio. If the test cannot make a decision
over a certain confidence threshold, then they increase the minibatch size and test again. This
process repeats until a decision.  The downside of this algorithm is the need to keep incrementing
the minibatch size; in the worst case, all $N$ data points may be needed in a single iteration.  A
similar approach from~\cite{icml2014c1_bardenet14} has slightly more robust theoretical properties,
but comes at the cost of more computational time.

@@

In this section, we explore the convergence properties of our MH test. Due to space constraints, all
our proofs are in Appendix~\ref{app:proofs}.

We first introduce some notation. Let our true and approximated acceptance probabilities be $P_a =
g(\Delta) = \Pr(\Delta + X > 0)$ and $P_a' = \Pr(\Delta' + X_{\rm corr} > 0)$.  Define $X_{\xi} =
\varepsilon - X_{\rm norm}$ which means, invoking Equations~\ref{eq:relationship}
and~\ref{eq:deconvolution}, that $X_\xi$ is the random variable characterizing the accuracy of our
$X_{\rm norm}$.  Thus, $\Delta' = \Delta + X_{\rm norm} + X_\xi$. We assume the standard deviation
of $X_{\rm norm}$ is less than $\pi/\sqrt{3}$, so that we can use Equation~\ref{eq:deconvolution}.
We denote $F_X(x)$ as the CDF of $X$.

@@

%% \begin{proof}
%% See Appendix~\ref{app:theory1}.
%% \end{proof}

@@

%% \begin{proof}
%% See Appendix~\ref{app:theory2}.
%% \end{proof}

@@

%% \begin{proof}
%% See Appendix~\ref{app:theory3}.
%% \end{proof}

@@

Figure~\ref{fig:gauss_mix_2} (to the left) shows a histogram of the final minibatch sizes used by
the adaptive subsampling method each iteration (capped at 600 for readability).  Usually, the method
can make a decision with 200 samples. Occasionally, however, it must use all 10000 points, resulting
in more computation (some of it wasted). For this particular experiment, that happened eight times.  In
contrast, our method keeps the minibatch size fixed at 200, but requires that ${\rm std}(\Delta') <
1.2$. Figure~\ref{fig:gauss_mix_2} (in the middle) shows a histogram of the estimated ${\rm
std}(\Delta')$ values each iteration. Only 75 iterations resulted in ${\rm std}(\Delta') \ge 1.2$
and for those, we skipped the rest of the iteration (setting $\theta_{t+1}=\theta_t$). To smooth the
${\rm std}(\Delta')$ values, we use moving average updates. The third plot in
Figure~\ref{fig:gauss_mix_2} investigates the number of times ${\rm std}(\Delta') \ge 1.2$. For six
minibatch sizes (50, 100, 150, 200, 250, and 300), we ran MCMC with our MH test five times and
averaged the number of violations. As minibatch sizes increases, the number of violations decreases,
showing that our test is more applicable.

@@

We next use logistic regression for the binary classification of 1s versus 7s in the MNIST
dataset~\cite{lecun-mnisthandwrittendigit-2010}. The data has 12007 and 1000 training and testing
points, respectively (we used a random subset of the test data). The propose is again a random walk
with covariance matrix $0.1I$ for the $784\times 784$ identity matrix $I$. We initialize the
posterior temperature at $T_0=3000$, and decrease it according to $T_i = T_0 /\sqrt{i + 1}$ for
iteration $i \ge 1$. We set the minibatch size $m=50$ and compare our minibatch MH test with the
adaptive minibatch.

Figure~\ref{fig:logistic_fig} shows the prediction accuracy and log likelihood on the test set as a
function of the cumulative training data points processed. Our test increments the cumulative data
by a fixed amount per iteration, but the adaptive method may require more data per iteration.  We
see that our minibatch MH test is more efficient; it has similar or better performance while
consuming fewer data points.

@@

We use this experiment to illustrate another property of our minibatch MH testing: when the size of
minibatch increase to a threshold, the accuracy / log likelihood of sampled data at a fix iteration
number will not improve with larger minibatch size, see Fig.~4c. This is because the variance of
$\Delta'$ decrease with large minibatch, and the error of our minibatch MH testing is proportional
to our estimation error of variance of $\Delta'$. When the ${\rm Var}(\Delta')$ decrease to a
threshold, the error of our minibatch MH testing can be ignored.

@@

In this experiment, we apply about minibatch MH testing with SGHMC (Stochastic Gradient Hamiltonian
Monte Carlo~\cite{sghmc_2014}) proposer to sample the parameter instances from a fully connected
neural network. We use Higgs data set from University of California Irvine. It is a classification
problem to distinguish between a signal process which produces Higgs bosons and a background process
which does not. We build a 4 layers neural network to tackle this problem. The first layer is a
input layer, the last layer is a softmax layer, the intermediate layers are linear followed by
Sigmoid, norm, dropout, starting and ending with linear operation.

@@

\begin{itemize}[noitemsep]
    \item We use $N=10000$ samples of $x_i$.
    \item The covariance of the Gaussian random walk is ${\rm diag}(0.1, 0.1)$.
    \item We use 10000 iterations. One iteration involves one $\theta'$ proposal and a test to
    accept/reject.
    \item The minibatch size is set at 100, so it is one percent of the total data. For the
    adaptive sampling approach, this is the amount that the minibatch size would keep increasing if
    we needed more precision. Our approach, of courses, keeps the size fixed.
    \item For adaptive sampling, the tolerance for deciding on a test is 0.05. The temperature for
    that is also set at $T=1$ because that algorithm was not designed to deal with temperatures.
    \item For our approach, to get an estimate of the standard deviation of $\Delta'$ (remember,
    $\Delta'$ is a minibatch approximation of $\Delta$) we take five other minibatches (randomly
    selected from the data) and take the standard deviation of those values.
\end{itemize}

@@

We applied our minibatch Metropolis Hastings algorithm to a Bayesian logistic regression model,
comparing with the adaptive sampling method. We tested both our method and the adaptive sampling
method using a random walk proposer $q(\theta \mid \theta_t) = N (\theta_t, \sigma^2)$. Since the
random walk proposer does not contain any information about the target distribution, it only relies
on the Metropolis-Hastings test to converge to the correct distribution. Therefore, random walk
proposer is the ideal choice for illustrating our algorithm and comparing with the adaptive sampling
method.
